Transcriptional dynamics of influenza virus infection in single cells

### 2017-09-08 paper/paper.tex
Added rough introduction. Changed one part of the methods that was not quite right. Have made some changes to Discussion, but still need to add paragraph on flu variation

\section{Introduction}

Heterogeneity is important in a lot of cellular processes even when isogenic~\citep{Shalek:2014ey,Shalek:2013ej}.

Population (genetic) heterogeneity is also important. 
Viral quasispecies, cancer single-cell, etc.
Salmonella paper (PhoP).

Literature on viral burst-size heterogeneity.
This goes back to Delbruck, Andino polio paper~\citep{Schulte:2014br}, the MDCK / flu paper.

Discuss segmented nature of influenza.
Maybe in the context of how this could further increase heterogeneity because there is a lot of potential for entire genes to be missing.
Includes Yewdell and Lowen papers.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_coinfection.pdf}}
\caption{\label{fig:coexpression}
The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
The bars show the logarithms of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
}
\figsupp[Co-infected cells express roughly equal amounts of a gene from each infecting viral variant. \label{figsupp:flowcyto}]{
{\bf(A)} Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were analyzed by flow cytometry for HA and eGFP expression.
{\bf(B)} The expression of HA and GFP are correlated in co-infected cells. 
Shown are the quantile-normalized HA and eGFP signals for double-positive cells. 
Cells are colored by density, using a Gaussian kernel density estimate. 
{\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
{\bf(F)} Gaussian kernel density estimate on HA positive cells from {\bf(C)}
}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D03.png}}
\figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
\figdata{The sequence of the HA viral RNA carrying the GFP gene is in \texttt{HAflank-eGFP.fasta}.\label{figdata:HAflank}}
\end{figure} 
Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
This observation is consistent with the finding by \citet{Dou:2017cp} and \citet{Huang:2008gy} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.

@@

\section{Discussion}
We have generated total transcriptome measurements of individual influenza-infected cells -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
Many of our observations closely match those made by other methods, validating our approach and granting us confidence in exploring more nuanced features of our data.
Overall we find that influenza replication is defined broadly by both significant stochasticity and significant robustness, and that a view of viral infection as either strictly deterministic or incredibly chaotic fails to accurately describe the infectious process.
Moreover we found that stress, likely induced by oxidative damage, appears to define the cellular response to infection.

One of the most striking features of stochasticity with respect to influenza infection is the absence of genes corresponding to a given genomic segment, observed both by ourselves and others~\citep{Brooke:2013kb,Dou:2017cp}
While there is significant speculation, and some indirect evidence, regarding the step of viral replication at which these genes are lost, there currently exists no strong consensus as to its identity -- indeed there may be no single event wholly responsible the stochastic loss of viral gene expression ~\citep{Brooke:2017gm, Heldt:2015iz}
Regardless it does not seem unreasonable to surmise that this is a consequence of possessing a segmented genome. If so, it can be reasonably concluded that at some phase of viral replication, whether packaging, delivery to the nucleus, production of replicative intermediates, or some as-yet unsurmised step, the components of the viral genome must act in at least a semi-independent fashion. 
A consequence of this stochasticity is a subpopulation of infected cells that express significantly lower levels of viral transcripts. 
Consistent with prior population-level studies of transcription, we find that absence of any of the four viral gene products necessary to form the viral polymerase leads to a dramatic drop in the levels of mRNA transcripts detected~\citep{Vreede:2004ip,eisfeld2015centre}. 
Our results are in slight contrast to a prior flow cytometric analysis of protein expression that described a relatively minor drop in expression associated with absence of viral ribonucleoprotein~\citep{Brooke:2013kb}.
This discrepancy could be explained by two possibilities. 
The first is that mRNA levels are not wholly predictive of viral protein abundance. 
The second is that flow cytometry depends on a user-defined dynamic range of sensitivity as well as positive and negative gates, and that the potential 2-3 log difference in population means was either not captured in the sensitivity of the instrument or that gates on positive populations excluded bone-fide positives. 
Indeed, gating on HA staining using a strict cutoff based on an uninfected population alone produces a curve that is consistent with a bimodal distribution of HA expression (Figure ~\ref{fig:coexpression} Figure Supplement ~\ref{figsupp:flowcyto}F).

@@

Perhaps even more important to influenza?s success in the face of stochasticity is the capacity of the virus to control the host innate immune response despite a large number of non-infectious events. 
While non-infectious events might be considered metabolic ?waste?, they would, theoretically, not impact the growth of successful viral infections. 
However, recognition of any virus, infectious or not, and production of interferon, would impact the entire viral population. 
While we cannot extrapolate much from our single interferon-positive cell, it can be stated that interferon induction by influenza is rare in our dataset. 
This is consistent with prior experiments using reporter cell lines,  which places the induction at ~1\% of cells with detectable influenza protein expression -- albeit this value is likely sensitive to differences in viral propagation and strain utilized ~\citep{Killip:2017ef, Chen:2010cr}. 
If the major influenza interferon antagonist, NS1, is absent in ~10-20\% of infections, and lowly expressed in those infections engaging in only primary transcription, we must posit that NS1 alone is insufficient to explain influenza evasion of innate immunity. 
Interestingly, if other segmented viruses also exhibit a pattern of stochastic segment loss, they too would need to overcome this hurdle and escape innate immunity through mechanisms relying on more than a gene product generated by a single segment.

@@

Wildtype A/WSN/1933 (H1N1) influenza virus was generated by reverse genetics using the plasmids pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
The sequences of the viral RNAs encoded in these plasmids are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
The synonymous mutations were chosen to introduce nucleotides that are common in naturally occurring H1N1 viruses to avoid unwanted attenuation.
The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}
To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at of the clarified viral supernatant -80 \si{\degreeCelsius }. 
We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours, when obvious cytopathic effect was noted.

>>>>

\section{Introduction}
Recent advances in sequencing and visualization technologies have revealed that cell-to-cell heterogeneity is a fundamental feature of biological processes -- impacting events from bacterial entrance into quiescent states to organogenesis in multicellular organisms~\citep{JohnstonJr:2010jn}.
Due to this heterogeneity, biological pathways generated from bulk population data may not accurately describe the timing, order, or even the identity of events within single cells ~\citep{Trapnell:2014kg}.
Nearly no other feature of biological life is as defined by heterogeneity as viral infection, wherein events can be highly stochastic due to the incredible combination of both genetic and non-genetic sources of variation.

The contribution of heterogeneity to infectious processes has been well appreciated in the study of viral quasispecies, wherein pathogenesis is best described as an emergent process dependent upon underlying genetic diversity ~\citep{Lauring:2010if}.
This concept has revolutionized the study of RNA viruses: including polio, hepatitis C, and influenza \citep{Echeverria:2015ew, Kuroda:2010gt}. 
Direct evidence of the contribution of the quasispecies to pathogenesis has been observed -- for instance mutant strains of polio that produce less variation exhibit significantly less fitness in animal models of infection despite displaying wild-type replication kinetics \emph{in vitro}~\citep{Vignuzzi:2005ju}.
However, these studies have largely been limited to population-level analyses, which can link standing diversity to complex biological phenomena but are unable to deconvolute precise events in individual infected cells.

Non-genetic variation has also been described to contribute to the host response to infection. 
Single-cell transcriptomic analyses have revealed incredibly coordinated responses driven by stochasticity, even in tissue culture models using genetically homogeneous cells.
This heterogeneity in response has been observed both with respect to pure ligands such as lipopolysaccharide, wherein differences must be host-cell intrinsic ~\citep{Shalek:2014ey,Shalek:2013ej,Rand:2012kl}, and with respect to pathogens, wherein differences are likely a complex mixture of host-cell intrinsic and pathogen-specific heterogeneity ~\citep{Avraham:2015cj}. 
Moreover, stochasticity is observed both in interferon production, which drives the anti-pathogen state, and the expression of the downstream effector molecules of innate immunity.
As inflammation is a potentially deleterious process, restricting a response to a small subset of cells upon initial recognition of a threat might be particularly advantageous.

Influenza, a segmented negative sense RNA virus, places a significant infectious burden on the global population and yet, despite many decades of study, there is still significant conjecture regarding those features of viral populations that drive observed phenotypes.
Cells infected with this virus exhibit a wide degree of heterogeneity, including a number of semi-infectious events wherein the full complement of viral gene products is absent~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}.
This has broadly been observed to impact rates of recombination, however its contribution to other viral behaviors remains unclear~\citep{Fonville:2015cg}. 
Influenza-infected cells are also observed to exhibit a wide-breadth of productivity -- exhibiting a 3-log range in infectious progeny even in high-MOI infections~\citep{Heldt:2015iz}. 
This appears to be a general trend of many RNA viral infections, from phage to polio, and the features underlying these observations appear multitudinous~\citep{DELBRUCK:1945wo, Schulte:2014br}.

We now hope to apply new single-cell transcriptomic technologies to the study of influenza replication in a tissue culture model of infection.
Early single-cell transcriptomic approaches were laborious -- requiring the generation of sequencing libraries on individual cells~\citep{Tang:2009kj,ld:2012gj}.
Breakthroughs in microfluidics helped to automate this process, but retained strictly parallel processing of sequencing libraries.
More recent technologies have introduced more complex means of barcoding and partitioning cells, allowing for the generation of libraries consisting of thousands of cells with ?single pot? downstream processing.
These include the use of reverse-phase emulsions, gravity capture, and even \emph{in situ} reverse transcription after cellular partitioning~\citep{JunSSi,Gierahn:2017bc,Klein:2015ki,Macosko:2015ht}. 
We hope to leverage these technological advances in the study of influenza, using transcriptomic analysis to understand the phenotypic heterogeneity that abounds during infection with this virus.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_coinfection.pdf}}
\caption{\label{fig:coexpression}
The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
The bars show the logarithms of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
}
\figsupp[Co-infected cells express roughly equal amounts of a gene from each infecting viral variant. \label{figsupp:flowcyto}]{
{\bf(A)} Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were analyzed by flow cytometry for HA and eGFP expression.
{\bf(B)} The expression of HA and GFP are correlated in co-infected cells. 
Shown are the quantile-normalized HA and eGFP signals for double-positive cells. 
Cells are colored by density, using a Gaussian kernel density estimate. 
{\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D02.png}}
\figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
\figdata{The sequence of the HA viral RNA carrying the GFP gene is in \texttt{HAflank-eGFP.fasta}.\label{figdata:HAflank}}
\end{figure} 
Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
This observation is consistent with the finding by \citet{Dou:2017cp} and \citet{Huang:2008gy} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.

@@

\section{Discussion}
We have generated total transcriptome measurements of individual influenza-infected cells -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
The dataset we have generated is largely consistent with prior flow cytometry-based analyses, allowing us to build upon those prior observations with a more holistic view, beginning to understand the consequences of heterogeneity in influenza.
Overall we find that influenza replication is defined broadly by both significant stochasticity and significant robustness, and that a view of viral infection as either strictly deterministic or incredibly chaotic fails to accurately describe the infectious process.
Moreover we found that stress, likely induced by oxidative damage, appears to define the cellular response to infection.

One of the most striking features of stochasticity with respect to influenza infection is the absence of genes corresponding to a given genomic segment, observed both by ourselves and others~\citep{Brooke:2013kb,Dou:2017cp}
While there is significant speculation, and some indirect evidence, regarding the step of viral replication at which these genes are lost, there currently exists no strong consensus as to its identity -- indeed there may be no single event wholly responsible the stochastic loss of viral gene expression ~\citep{Brooke:2017gm, Heldt:2015iz}
Regardless it does not seem unreasonable to surmise that this is a consequence of possessing a segmented genome. If so, it can be reasonably concluded that at some phase of viral replication, whether packaging, delivery to the nucleus, production of replicative intermediates, or some as-yet unsurmised step, the components of the viral genome must act in at least a semi-independent fashion.

A consequence of this stochasticity is a subpopulation of infected cells that express significantly lower levels of viral transcripts. 
Consistent with prior population-level studies of transcription, we find that absence of any of the four viral gene products necessary to form the viral polymerase leads to a dramatic drop in the levels of mRNA transcripts detected~\citep{Vreede:2004ip,eisfeld2015centre}. 
Our results are in slight contrast to a prior flow cytometric analysis of protein expression that described a relatively minor drop in expression associated with absence of viral ribonucleoprotein~\citep{Brooke:2013kb}.
This discrepancy could be explained by two possibilities: the first is that mRNA measurements may not be wholly reflective of protein levels, and the second is that flow cytometry requires decisions regarding sensor sensitivity and positive gates and the potential biases that those decisions incur.
While sequencing-based analysis comes with its own set of caveats, the detection threshold is 'baked-in' and so potentially less subject to bias.

@@

Perhaps even more important to influenza's success in the face of stochasticity is the capacity of the virus to control the host innate immune response despite a large number of non-infectious events. 
While non-infectious events might be considered metabolic 'waste', they would, theoretically, not impact the growth of successful viral infections. 
However, recognition of any virus, infectious or not, and production of interferon, would impact the entire viral population. 
While we cannot extrapolate much from our single interferon-positive cell, it can be stated that interferon induction by influenza is rare in our dataset. 
This is consistent with prior experiments using reporter cell lines,  which places the induction at ~1\% of cells with detectable influenza protein expression -- albeit this value is likely sensitive to differences in viral propagation and strain utilized ~\citep{Killip:2017ef, Chen:2010cr}. 
If the major influenza interferon antagonist, NS1, is absent in ~10-20\% of infections, and lowly expressed in those infections engaging in only primary transcription, we must posit that NS1 alone is insufficient to explain influenza evasion of innate immunity. 
Interestingly, if other segmented viruses also exhibit a pattern of stochastic segment loss, they too would need to overcome this hurdle and escape innate immunity through mechanisms relying on more than a gene product generated by a single segment.

@@

Wildtype A/WSN/1933 (H1N1) influenza virus was generated by reverse genetics using the plasmids pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
The sequences of the viral RNAs encoded in these plasmids are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}
To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at of the clarified viral supernatant -80 \si{\degreeCelsius }. 
We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours, when obvious cytopathic effect was noted.
